PanGraph: scalable bacterial pan-genome graph construction \\ Supplementary Information

### 2023-02-18 reviewer_response.tex

      The software was easy to install and the documentation available online is impressive.

      There was no comparison with other tools - would be appropriate if e.g. downstream mapping performance were evaluated but that is outside the scope here.\\

      On reading I had a few queries, which could perhaps be clarified further in the text.}

\reviewer{R1.1}{A pancontig is defined as a multiple sequence (sub)alignment. In line 54 you say "Pairwise graph alignment is performed by an all-to-all alignment of the pancontigs" which I interpreted to mean all sequences in the alignment vs all sequences in another alignment. You do later add that "To align two graphs, the consensus sequences of all pancontigs in both graphs are searched for homologies and aligned". Adding the word consensus in the earlier sentence might avoid confusion.}
% \Marco{changed to ``Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} consensus sequences between both graphs, and the order of pairwise alignments is determined by a guide tree." Too compressed? I could also break in two sentences and expand a bit more.}\\
% \Liam{that's clear to me.}
\response{We agree with the reviewer suggestion. The wording in Line 54 has been updated to ``all-to-all alignment of the \textit{pancontigs} consensus sequences''.}

\reviewer{R1.2}{Figure 4C: The main text talks about how to interpret this, as ~1/N for TB and a bit less compression for E. coli and K. pneumoniae where the core is a smaller fraction, but this is not so easy to interpret from the figure. Would it be possible to rescale this plot to orders of magnitude of N? Or in another way aid this reading?}
\response{We agree with the reviewer that it is important to be able visually compare the compression with the number $N$ of isolates for each species, and we thank him for the suggestion. We modified fig.~4C in the main text and fig.~4I in SI marking the value of $1/N$ for each species with horizontal lines, and updated the captions accordingly.}
% \Marco{maybe try visualizing it as deviation from "one strain" compression (1/N)? But the "maximal" compression depends not only on the sample size N but also on the gene frequency distribution. Another possible comparison might be to pangraph's estimation of the pangenome size, but again we might have different effects coming into play.}\\
% \Liam{yes, I think just adding a dashed line for 1/N for each species to the plot would aid that reading and satisfy this point.}

\reviewer{R1.3}{Figure 5 has 2 typos: pacontigs and panconting.}
% \Marco{fixed.}
\response{The typos have been corrected.}

\reviewer{R1.4}{Throughout the text E. Coli and K. Pneumoniae etc capitalization.}
% \Marco{fixed.}
\response{The capitalization has been corrected.}

@@

\reviewer{R3}{An issue I have struggled with in my own work is how to represent an orthologous gene cluster (node/vertex) within the PGG when the gene sequence in some genomes has an insertion event such as an IS element within the gene. I assume for PanGraph that this would necessitate splitting the node into multiple pancontigs. How is this sustainable for species with large numbers of active IS elements? This could also be investigated for PanGrah using suitable simulated data.}
\Liam{I think nothing to say here apart from `yes, active IS elements split pancontigs up' (and can themselves be split up due to similarity with other IS\ldots). For the future one might want to have pangenome methods where you can explicitly pass a list of predefined sequences that should not be split. But I think this is a general problem and not something to investigate with simulations or fix here.}\\
\Marco{agree. I think this is a structural property that we want to capture, the fact that these pancontigs are split.}

\bibliography{cite}{}

\end{document}

% meeting notes:
% SIAS -> (new) Supplementary Information Algorithm Section
% make changes in the text traceable in the reviewer answers.
% Reviewer 1:
% - for blocks, mention that consensus is considered as the centroid of a star-shaped phylogeny?
% - 2) add line to fig. on compression
% - 5) the effect comes from the fact that core genes are in long syntenic stretches, compared to accessory that are often in messy islands (cite Rocha?). Just add half sentence to the paper.
% - 6) annotate the figure with numbers
% - 8) reference: graph alighner, ggCaller? Many take GFA as input.\
% - 10) this is something that could be explained in the SIAS: one starts with long and shallow blocks and as graphs are merged and propagated down the tree the blocks become shorter and deeper. Removing energy threshold just causes more merging and deeper but shorter blocks. (figure?)
% - test pangraph with 500 E.coli.
% Reviewer 2:
% - 1) do not mention blocks, only pancontigs
% - specify better the meaning of null energy parameters and why more merging.
% Reviewer 3:
% - paralog merging is partially controlled by beta. As future direction one should build trees for every blok and split paralogs in a PanX fashion. 
% - more nuanced treatment of the literature.
% - update marginalize figure legend
% - update SI section numbering in the main text

### 2023-02-18 supplementary.tex
feat: review comments
Update on Overleaf.
chore: partial update supplementary

\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\Lcore}{L_\text{core}}
\newcommand{\Lpang}{L_\text{pang}}
\newcommand{\Lgen}{L_\text{gen}}
\newcommand{\dcore}{\langle d_\text{core} \rangle}

@@

@@

\begin{figure}[htb]
    \includegraphics[width=\textwidth]{figs_suppl/panx_benchmark_suppl.pdf}
    \caption{{\bf Benchmark on real data}. We run \textit{PanGraph} on datasets consisting of 5 different bacterial species, and using 5 different options for the alignment kernel as described in the text. Runs were performed using 8 cores.
        \textbf{A}: number of isolates for each species.
        \textbf{B}: algorithm wall-time.
        \textbf{C}: maximum memory usage.
        \textbf{D}: total number of pancontigs in the pangenome graph.
        \textbf{E}: minimum number of pancontigs that can add up to 50\% of the total pangenome graph length.
        \textbf{F}: threshold length such that the cumulative size of all pancontigs longer than this threshold is more than 50\% of the full pangenome graph.
        \textbf{G}: total length of the pangenome graph (sum of all pancontig lengths).
        \textbf{H}: fraction of the pangenome graph that is composed of core pancontigs, i.e. pancontigs that occur exactly once per isolate.
        \textbf{I}: ratio between the length of the pangenome graph and the total length of all genomes included in the graph.
    }
    \label{fig:panx-benchmark-suppl}
\end{figure}

>>>>

\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\Lcore}{L_\text{core}}
\newcommand{\Lpang}{L_\text{pang}}
\newcommand{\Lgen}{L_\text{gen}}
\newcommand{\dcore}{\langle d_\text{core} \rangle}
\newcommand{\Lthr}{L_{\min}}

@@

\section{Graph merging algorithm}
\label{supp:graph_merging}

In this section we explain in more detail the procedure used to merge two pangenome graphs together. This procedure is at the core of PanGraphs's algorithm.

As explained in the main text, the full pangenome graph is built by placing genomes on terminal leaves of a guide tree. These genomes are initially considered as elementary graphs composed of a single \textit{pancontig}. These graphs are then propagated along the tree towards the root, and at each intersection a merge operation is performed. The full pangenome graph is then collected from the root.

\subsection*{pangenome graph representation}

As a preliminary step in the explanation we spend few words on the pangenome graph representation of pangraph. A pangenome graph is essentially composed of two elements: pancontigs, which encode for alignment of homologous sequences, and paths, that correspond to representation for genomes as a sequence of pancontigs.

Each \textbf{pancontig} contains the full information on the alignment of homologous sequences it contains. In PanGraph each pancontig is econded with a consensus sequence, and each line of the alignment is saved as variations over the consensus (mutations, insertions and deletions). More information on the precise encoding used can be found in PanGraph documentation.\footnote{See \url{https://neherlab.github.io/pangraph/tutorials/tutorial_2/\#From-block-consensus-to-node-alignment}}

Each \textbf{path} is an oriented collection of pancontigs that represents a genome. For each item in the sequence along with the pacontig id PanGraph also stores the pancontig orientation (whether homology with the consensus is on the forward the reverse strain of the genome) and which variations on the pancontig consesus (i.e. which line in the pancontig alignment) correspond to this precise location in the genome.

From knowledge of all the pancontigs and pahts, every single genome can be exactly reconstructed. Every part of the sequence is always encoded in a pancontig. ``Small-scale'' difference between genomes (i.e. SNPs, small insertions and deletions) are encoded in pancontigs as variations over the consensus. ``Large-scale'' differences, such as presence or absence of long stretch of sequences, are instead encoded in paths as presence or absence of pancontigs. This scale separation can be tuned using PanGraph parameters, as will be explained in the next section.

\subsection*{graph merging}

At each merge operation two pangenome graphs need to be compressed into a single one. Three of pangraph's parameters play a role in the outcome of this procedure:
\begin{itemize}
    \item the minimal block length $\Lthr$.
    \item pseudo-energy parameters $\alpha$ and $\beta$ (cf. eq.~1 in the main text).
    \item the choice of alignment kernel (minimap2 or mmseqs) and aligner precision (asm5, asm10 or asm20 for minimap2).
\end{itemize}
In the merging, these parameters control how much of the diversity between sequence must be stored as large-scare variation (difference in paths) and how much as fine-grained variation (differences in alignments, stored in pancontigs). This is done for example by setting a divergence limit for sequences to be stored in the same alignment.\\
An implicit assumption of our procedure is that the phylogeny of sequences stored inside the same panconting is roughly star-like, with each sequence containing independent variations from the consensus. In such cases the consensus is a good ``average'' representation for the set of sequences in the alignment.
In the next section we briefly discussion how PanGraph behaves when this assumption is not satisfied, such as in the case of merging ortholog and paralog sequences in the same pancontig.\\
% move this paragraph or change?

The merging operation is performed in the following steps:
\begin{enumerate}
    \item Using the chosen aligner kernel, we scan for homology between pancontigs of the two different graphs. This is done by using only the consensus sequence of each pancontig.
    \item The aligner will return a set of matches. Each match is characterized by an alignment length and a measure of diversity. Using this information we rank matches by pseudo energy, so that longer and less diverged matches will be processed first. We refuse all matches with negative pseudo-energu and all matches shorter than the threshold length $\Lthr$.
    \item For each of the remaining matches we attempt to merge pancontigs following these steps:
          \begin{enumerate}
              \item We check that no prior merges have been performed on the same two pancontigs. Sometimes multiple matches are found in the same pancontig, and in this case we prioritize the ones with higher pseudo-energy. Any following match will be discarded in this step. The same merge might however be performed later on in the self-map phase of the algorithm (see step 4).
              \item Flanking regions of the consensus sequences that are not part of the matches are cut and not merged and stored in separate pancontigs. One exceptions is when one of the flanking regions is shorted than the threshold length $\Lthr$. In this case instead of creating a very short pancontig this region is stored as a small flanking \textbf{insertion (check)} in the merged pancontig.
              \item The CIGAR string of the match is then parsed and subdivided in sub-regions. Regions can be either shared or private to query or reference. Shared regions contain Each insertion or deletion longer than $\Lthr$ will cause a split in the merge, and will be stored as a separate panconting. \textbf{(what happens with small matches in longer flanking indels?)}
              \item
          \end{enumerate}
    \item Once the list of all matches has been exhausted and all merges have been performed, the two graphs have been turned into
\end{enumerate}

% post-processing step: transitive edges. Explain.

% blocks become progressively deeper but shorter. The energy parameters and the threshold lengt determine when to stop.
% implicit assumption of star-like topology.

\subsection*{PanGraph parameters effect}
% scale separation effect. Maybe move somewhere else?
% also discussion on orthologs and paralogs.

@@

\begin{figure}[htb]
    \includegraphics[width=\textwidth]{figs_suppl/panx_benchmark_suppl.pdf}
    \caption{{\bf Benchmark on real data}. We run \textit{PanGraph} on datasets consisting of 5 different bacterial species, and using 5 different options for the alignment kernel as described in the text. Runs were performed using 8 cores.
        \textbf{A}: number of isolates for each species.
        \textbf{B}: algorithm wall-time.
        \textbf{C}: maximum memory usage.
        \textbf{D}: total number of pancontigs in the pangenome graph.
        \textbf{E}: minimum number of pancontigs that can add up to 50\% of the total pangenome graph length.
        \textbf{F}: threshold length such that the cumulative size of all pancontigs longer than this threshold is more than 50\% of the full pangenome graph.
        \textbf{G}: total length of the pangenome graph (sum of all pancontig lengths).
        \textbf{H}: fraction of the pangenome graph that is composed of core pancontigs, i.e. pancontigs that occur exactly once per isolate.
        \textbf{I}: ratio between the length of the pangenome graph and the total length of all genomes included in the graph. Since maximal compression also depends on the number of isolates $N$, we mark with horizontal lines the value of $1/N$ for each species.
    }
    \label{fig:panx-benchmark-suppl}
\end{figure}
